

Seminar 5 
========================================================

```{r}
library(lattice) 
```


```{r}
prDat <- read.table("C:/Users/Kateryna/Documents/RStudio/GSE4051_data.tsv")
str(prDat, max.level = 0)
prDes <- readRDS("C:/Users/Kateryna/Documents/RStudio/GSE4051_design.rds")
str(prDes)
````

###############

Take the Affymetrix probeset ID(s) as input and create a dataframe

```{r}
prepareData <- function(keepGenes) {
  miniDat <- subset(prDat, rownames(prDat) %in% keepGenes)
  # reshape the data to be tall and skinny
  miniDat <- data.frame(gExp = as.vector(t(as.matrix(miniDat))),
                        gene = factor(rep(rownames(miniDat), each = ncol(miniDat)),
                                      levels = keepGenes))
  miniDat <- suppressWarnings(data.frame(prDes, miniDat))
  return(miniDat)
}
```


Test a function to create a dataset for a small set of genes
```{r}
jDat <- prepareData(c("1419655_at","1438815_at"))
str(jDat)
head(jDat)
tail(jDat)
stripplot(gExp ~ devStage | gene, jDat,
          group = gType, jitter.data = TRUE,
          auto.key = TRUE, type = c('p', 'a'), grid = TRUE)
```


###############

A function to stripplot a mini-dataset
```{r}
makeStripplot <- function(df) {
  stripplot(gExp ~ devStage | gene, df,
            group = gType, jitter.data = TRUE,
            auto.key = TRUE, type = c('p', 'a'), grid = TRUE)
}
```


Test the stripplot function
```{r}
makeStripplot(jDat)
makeStripplot(newDat <- prepareData("1456341_a_at"))
str(newDat)
head(newDat)
```


###############

Test for a difference in expected gene expression for probeset "1456341_a_at" at developmental stage P2 vs. 4 weeks post-natal (ignoring genotype, i.e. lump the wild types and knockouts together).


Assume a common variance in the two groups.
```{r}
ttInput <- subset(prepareData("1456341_a_at"), devStage %in% c("P2","4_weeks"))
t.test(gExp ~ devStage, ttInput)
```


###############

Fit a linear model with a categorical covariate = do "one-way ANOVA". Focus on probeset "1438786_a_at".
```{r}
makeStripplot(mDat <- prepareData("1438786_a_at"))
```

Let's focus just on the wild type data for now. 
Model expression as a function of the devStage factor.
```{r}
mFit <- lm(formula = gExp ~ devStage, data = mDat, subset = gType == "wt")
summary(mFit)
```


Vet your inferential results: does the intercept look plausible given the plot? How about the devStageP2 effect, etc.?
_Yes, looks right._

###############

Test whether the P2 and P10 effects are equal or, equivalently, whether their difference is equal to zero.
Construct the contrast matrix to form the difference between the P2 and P10 effects.
```{r}
contMat = matrix(c(0, 1, 0, -1, 0), nrow=1)
(obsDiff <- contMat %*% coef(mFit))
```

Check that this really is the observed difference in sample mean for the wild type mice, P2 vs. P10.
```{r}
(sampMeans <- aggregate(gExp ~ devStage, mDat, FUN = mean,
                        subset = gType == "wt"))
with(sampMeans, gExp[devStage == "P2"] - gExp[devStage == "P10"])
```


Calculate the (estimated) standard error for our contrast. The variance-covariance matrix can be obtained with vcov():
```{r}
vcov(mFit)
```

Let's check that this is really true. If we take the diagonal elements and take their square root, 
They should be exactly equal to the standard errors reported for out original model.
```{r}
summary(mFit)$coefficients[ , "Std. Error"]
sqrt(diag(vcov(mFit)))
```


A typical matrix of inferential results 
```{r}
summary(mFit)$coefficients
```


Variance-covariance matrix of a contrast 
```{r}
(estSe <- contMat %*% vcov(mFit) %*% t(contMat))
```

Form a test statistic as an observed effect divided by its estimated standard error:
```{r}
(testStat <- obsDiff/estSe)
```


Compute a two-sided p-value 
```{r}
2 * pt(abs(testStat), df = df.residual(mFit), lower.tail = FALSE)
```

Not surprisingly, this p-value is rather large and we conclude there is no difference.

##################
Fit a linear model with two categorical covariates

Focus on probeset "1448690_at"
```{r}
makeStripplot(oDat <- prepareData("1448690_at"))
str(oDat)
```


Fit a linear model with covariates gType and devStage and include their interactions.
```{r}
oFitBig <- lm(formula = gExp ~ gType * devStage, data=oDat)
summary(oFitBig)$coef
```
Vet the results. Is the intercept plausible? How about the various effects? Do the ones with small p-values, e.g. meeting a conventional cut-off of 0.05, look 'real' to you?
_Intercept and effects seem reasonable_

Fit a related, smaller model with the same covariates, but this time omit the interaction.
```{r}
oFitSmall <- lm(formula = gExp ~ gType + devStage, data=oDat)
```

Let's determine if the interaction terms are truly necessary. Recall the anova() function can take two fitted models, 
one nested within the other, and perform this test.
```{R}
anova(oFitSmall, oFitBig)
```

With a p-value very close to one, we confirm that there is no evidence for interaction in this particular case.

Looking at a more interesting gene
```{r}
makeStripplot(oDat2 <- prepareData("1429225_at"))
oFitBig2 <- lm(formula = gExp ~ gType * devStage, data=oDat2)
oFitSmall2 <- lm(formula = gExp ~ gType + devStage, data=oDat2)
anova(oFitSmall2, oFitBig2)
```

The interaction here is highly statistically significant.

################
Further work
use data aggregation strategies from last week to do some of the same work for small sets of genes
```{r}
genes = c("1419655_at","1438815_at")
jDat <- prepareData(genes)
library(plyr)
(aggr <- ddply(jDat, ~ sidChar + devStage + gType, summarize, gExp = mean(gExp), gene = paste(genes, collapse=' & ')))
makeStripplot(aggr)
```


Fit linear and quadratic models to the expression data for one or several genes
```{r}
library(car)
mDat <- prepareData("1438786_a_at")
mDat$age <- recode(mDat$devStage, "'E16'=-2; 'P2'=2; 'P6'=6; 'P10'=10; '4_weeks'=28", as.factor.result = FALSE)
head(mDat)
library(lattice)
```

Linear fit
```{r}
linFit = lm(formula=gExp~age, data=mDat)
summary(linFit)
```

Draw points and a LOESS fitted regression curve. Other possible types: "p", "l", "h", "b", "o", "s", "S", "r", "a", "g", "smooth", "spline"
Docs: https://stat.ethz.ch/R-manual/R-devel/library/lattice/html/panel.xyplot.html
```{r}
xyplot(gExp ~ age, mDat, type=c("smooth", "p"))
```


Quadratic fit
```{r}
quadFit = lm(formula=gExp~age + I(age^2), data=mDat)
summary(quadFit)
xyplot(gExp ~ age^2, mDat, type=c("smooth", "p"))
``` 

4 week developmental stage generally posed a difficult fitting problem for the quadratic model where we regressed expression on age. Шt is simply too far separated in time to be easily modelled quantitatively with the other 4 developmental stages.


Let's drop the 4 week data and revisit this dataset with linear and quadratic models.
```{r}
(mDatSubset <- subset(mDat, devStage != '4_weeks'))
mDatSubset$age <- recode(mDatSubset$devStage, "'E16'=-2; 'P2'=2; 'P6'=6; 'P10'=10", as.factor.result = FALSE)
linFit = lm(formula=gExp~age, data=mDat)
summary(linFit)
xyplot(gExp ~ age, mDatSubset, type=c("smooth", "p"))
```

```{r}
quadFit = lm(formula=gExp~age + I(age^2), data=mDat)
summary(quadFit)
xyplot(gExp ~ age^2, mDatSubset, type=c("smooth", "p"))
```


