Seminar 4 
======================


```{r}
library(lattice)
library(plyr)
```

Load the data 
```{r}
prDat <- read.table("C:/Users/Kateryna/Documents/RStudio/GSE4051_data.tsv")
```
```{r}
head(prDat)
str(prDat, max.level = 0)
prDes <- readRDS("C:/Users/Kateryna/Documents/RStudio/GSE4051_design.rds")
str(prDes)
```


Extract the data for one gene
```{r}
set.seed(987)
(theGene <- sample(1:nrow(prDat), 1))
pDat <- data.frame(prDes, gExp = unlist(prDat[theGene, ]))
str(pDat)
```


```{r}
aggregate(gExp ~ gType, pDat, FUN = mean)
stripplot(gType ~ gExp, pDat)
```

two-sample t test comparing wild type to the Nrl knockouts
```{r}
ttRes <- t.test(gExp ~ gType, pDat)
str(ttRes)
ttRes$statistic
ttRes$p.value
```

Try another gene selected at random
```{r}
(theGene2 <- sample(1:nrow(prDat), 1))
pDat2 <- data.frame(prDes, gExp = unlist(prDat[theGene2, ]))
aggregate(gExp ~ gType, pDat2, FUN = mean)
stripplot(gType ~ gExp, pDat2)
```

Do t-test, Wilcox and KS tests 
```{r}
ttRes2 <- t.test(gExp ~ gType, pDat2, var.equal=FALSE)
ttRes2cv <- t.test(gExp ~ gType, pDat2, var.equal=TRUE)
wRes <- wilcox.test(gExp ~ gType, pDat2)
ksRes <- ks.test(pDat2$gType, pDat2$gExp)

ksRes <- suppressWarnings(ks.test(pDat2$gType, pDat2$gExp))
```

Create a dataframe for tests results
```{r}
data.frame("method"=c(ttRes2$method, ttRes2cv$method, wRes$method, ksRes$method), 
           p.value=c(ttRes2$p.value, ttRes2cv$p.value, wRes$p.value, ksRes$p.value))
```


Load the  gene expression  dataset


```{r}
kDat <- read.table("C:/Users/Kateryna/Documents/RStudio/GSE4051_MINI.txt")
str(kDat)
kMat <- as.matrix(kDat[c('crabHammer', 'eggBomb', 'poisonFang')])
str(kMat)
```

Ways to get the median expression (in apply function 2 means columns) for a column
```{r}
median(kMat[ , 1])
median(kMat[ , "crabHammer"])
apply(kMat, 2, median)
apply(kMat, 2, quantile, probs = 0.5)


apply(kMat, 2, quantile, probs = c(0.25, 0.75))
```

minimum gene expression for each sample
```{r}
apply(kMat, 1, min)
colnames(kMat)[apply(kMat, 1, which.min)]
```

Computing row-wise sums
```{r}
rowSums(kMat)
```

do the rowSums and apply functions give the same results?
```{r}
all.equal (rowSums(kMat), apply(kMat, 1, sum))
```

Average expression of eggBomb for different levels of devStage

```{r}
aggregate(eggBomb ~ devStage, kDat, FUN = mean)
aggregate(eggBomb ~ gType * devStage, kDat, FUN = mean)
```


...or using plyr
```{r}
ddply(kDat, ~ devStage, summarize, exp = mean(eggBomb))
```


Pick 6 genes: 3 are interesting, 3 are boring

```{r}
keepGenes <- c("1431708_a_at", "1424336_at", "1454696_at",
               "1416119_at", "1432141_x_at", "1429226_at" )

miniDat <- subset(prDat, rownames(prDat) %in% keepGenes)
```

Data reshaping 


```{r}
miniDat <- data.frame(gExp = as.vector(t(as.matrix(miniDat))), gene = factor(rep(rownames(miniDat), each = ncol(miniDat)), levels = keepGenes))
```

          
```{r}
miniDat2 <- data.frame(prDes, miniDat)
head(miniDat2)
miniDat2 <- suppressWarnings(data.frame(prDes, miniDat))
```

Plot  3 'hits' and 3  boring genes
```{r}
stripplot(gType ~ gExp | gene, miniDat2, scales = list(x = list(relation = "free")),group = gType, auto.key = TRUE)
``` 
          
Two group comparisons for each of these 6 genes
          
```{r}
someDat <- droplevels(subset(miniDat2, gene == keepGenes[1]))
          t.test(gExp ~ gType, someDat)
```

==============================
Take-home work

Pick 100 genes at random
```{r}
genesSet <-  sample(1:nrow(prDat), 100)

dataset <- rownames(prDat)[genesSet]

q <- subset(prDat, rownames(prDat) %in% dataset)

z <- data.frame(gExp = as.vector(t(as.matrix(q))),
                gene = factor(rep(rownames(q), each = ncol(q)),
                              levels = dataset))
head (z)
y <- suppressWarnings(data.frame(prDes, z))
ttRes <- ddply(y, ~ gene, function(z) {
  zz <- t.test(gExp ~ gType, z)
  zy <- suppressWarnings(wilcox.test(gExp ~ gType, z))
  round(c(pVal_ttest = zz$p.value, pVal_W = zy$p.value ), 4)
})
head(ttRes)
```

